---
output: github_document
---


```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

<!-- badges: start -->
[![Travis build status](https://travis-ci.com/gdauby/ConR.svg?branch=master)](https://travis-ci.com/gdauby/ConR)
<!-- badges: end -->
[![CRAN RStudio mirror downloads](https://cranlogs.r-pkg.org/badges/ConR)](https://www.r-pkg.org/pkg/ConR)



The ideas behind this package, and its testing involved many people and institutes.
See the original paper published in [Ecology and Evolution](https://onlinelibrary.wiley.com/doi/full/10.1002/ece3.3704).



## Install R, ConR and dependent packages

**First step**. Install [R](https://cran.r-project.org/).

**Second step**. A proper way to work with R is to define a [working directory](https://bookdown.org/ndphillips/YaRrr/the-working-directory.html). If you are working with Rstudio, you can create the first time a [project](https://bookdown.org/ndphillips/YaRrr/projects-in-rstudio.html), which much simplify handling of scripts and data.


**Third step**. Install ConR package (one of the two following) :

To install the version on CRAN
```
install.packages("ConR")
```

To install the version under development.
```
install.packages("devtools")
devtools::install_github("gdauby/ConR")
```


# Loading ConR


**Attach ConR package**
This should be done everytime you open an R session.
```
library(ConR)
```
```{r loadData,include=FALSE}
library(ConR)
# source("D:/MonDossierR/ConR_orig/R/IUCNeval.functionv11.R")
data(dataset.ex)
# load("D:/MonDossierR/ConR_orig/data/dataset.ex.RData")
MyData <- dataset.ex
MyData <- MyData[which(!MyData[,3] %in% c("species_1", "species_2")),]
MyData$tax <- as.character(MyData$tax)
```

**Help files**
Any function in R is documented by a help file which can be obtained by the following code:
```
?EOO.computing
?IUCN.eval
?map.res
?subpop.comp
```




## Funding

The development of this package was supported by: 

  - the European Union’s Horizon 2020 research and innovation program under the Marie Skłodowska-Curie grant agreement No 795114.
  
  - CESAB (Centre for the Synthesis and Analysis of Biodiversity). Research program of the FRB (Foundation for Research on Biodiversity) under the [RAINBIO project](https://gdauby.github.io/rainbio/index.html).

## Acknowledgements


## See Also

Other R packages related to IUCN assessments:

  - `rredlist` (https://cran.r-project.org/web/packages/rredlist/rredlist.pdf)
  
  - `red` (https://cran.r-project.org/web/packages/red/red.pdf)
  
  - `redlistr` (https://cran.r-project.org/web/packages/redlistr/redlistr.pdf)





