---
title: "Conservation assessments based on multiple-IUCN criteria using ConR"

date: "`r format(Sys.time(), '%B %Y')`"
author: 
  - Renato A. F. de Lima^[Universidade de São Paulo, https://github.com/LimaRAF] and Gilles Dauby^[IRD, https://github.com/gdauby]
output:
    rmarkdown::html_vignette:
        toc: true
        number_sections: true
    md_document:
      variant: gfm

vignette: >
  %\VignetteIndexEntry{Introduction to ConR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{rmarkdown::render}
    \usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", echo = TRUE)
```

```{r load, echo = FALSE, eval = TRUE, include=FALSE}
devtools::load_all() # for development
```

<br/><br/>

This vignette show how to use ``ConR`` to apply a workflow for conducting
conservation assessments for multiple species using the IUCN criteria A, B, C
and D. Here, we use only three species to make it simple. But ``ConR`` was
developed to perform assessments for many more species at the same time.

<br/><br/>

# Definitions

We will use as examples three emblematic and threatened arborescent species that
are endemic or near-endemic to the Atlantic Forest in South America: *Araucaria angustifolia*, *Euterpe edulis* and *Paubrasilia echinata*. Thus, it is
necessary to obtain information for this specific part of the world (e.g.
habitat cover maps, protected areas). Here, we assume that this information is
already available.

```{r species, eval = TRUE, collapse = TRUE}
spp <- c("Araucaria angustifolia", "Euterpe edulis", "Paubrasilia echinata")
```


<br/><br/>

# Species information

To perform the IUCN conservation assessments we will need different types of
species information, related to their biology, distribution, population size,
threats, among other information.

## Occurrence records 

First, we will load some real occurrence data for these species. This example
data is provided with ``ConR`` in the internal object `example_tutorial`, but
users can load a file containing a similar data frame. We will save it as `occs`
for simplicity.

```{r occurrences, eval = TRUE, collapse = TRUE}
data(example_tutorial)
occs <- example_tutorial
head(occs, 2)
```

## Species biology and uses

Then, we need to define objects containing basic species information that are
necessary for the assessments.

```{r species-info, eval = TRUE, collapse = TRUE}
gen.length <- c(40, 65, 50) ## generation length in years
p.mature <- c(0.4485, 0.3028, 0.4143) ## TO DEFINE
early.sucession <- c(0.9, 1, 1) ## TO DEFINE
harvest <- c(10, 10, 10) ## TO DEFINE
```

## Population sizes

```{r pop-sizes, eval = TRUE, collapse = TRUE}
araucaria <- c(2263714973, 1763953733, 1742714122, 1706724090, 1672711025, 
               1648088967, 1597720526, 1571864195, 1500976994, 1406079272, 
               1283024003, 1212589860, 1196021554, 1115185361, 1142419430, 
               1163973686, 1159640510, 1162421409)
euterpe <- c(4131301664, 3423126071, 3375255741, 3301309873, 3173853514, 
             3060527245, 2844478840, 2668655717, 2457462150, 2292870303, 
             2032409445, 1726153167, 1713775781, 1695875046, 1691503991, 
             1700665970, 1696332109, 1702100498)
paubrasil <- c(24777189, 20519325, 20094366, 19379164, 18179102, 17154073, 
               15487787, 14304065, 12087519, 11100112, 8456852, 5808083, 
               5800533, 5970905, 6115215, 6181785, 6190677, 6268826)
years <- c("1850", "1940", "1945", "1950", "1955", "1960", "1965", "1970", 
           "1975", "1980", "1985", "1992", "1995", "2000", "2005", "2010", 
           "2015", "2018")
pop.sizes <- rbind(araucaria, euterpe, paubrasil)
colnames(pop.sizes) <- years
row.names(pop.sizes) <- spp
```


```{r est-pop-sizes, eval = TRUE, collapse = TRUE}
missing.years <- c(1853, 1858, 1868, 1878, 1883, 1888, 1898, 1908, 1913, 1918, 
                   1928, 1938, 1943, 1948, 1953, 1958, 1963, 1968, 1973, 1978, 
                   1983, 1988)
results <- vector("list", length(spp))
for (i in seq_along(spp)) {
  res <- pop.decline.fit(pop.size = pop.sizes[i,],
                  years = years,
                  project.years = missing.years,
                  models = c("quadratic", "exponential", 
                             "logistic", "general_logistic"), 
                  plot.fit = FALSE)
  results[[i]] <- res$data[,c("Year", "Observed", "Predicted")]
}

years <- results[[1]]$Year
ncols <- length(years)
nrows <- length(spp)
mean.pop <- matrix(NA, ncol = ncols, nrow = nrows, dimnames = list(spp, years))
for(i in seq_along(spp)) mean.pop[i,] <- results[[i]]$Predicted
```

 

<br/><br/>

# Population metrics

<br/><br/>

## Species distribution metrics 

First we need to load ``ConR`` to make the functions available in your R
workspace.

```{r setup}
library(ConR)
```

### Extent of Occurrence (EOO)
```{r eoo, eval = TRUE, collapse = TRUE}
EOO.hull <- EOO.computing(XY = occs[, c(1:3)], method.range = "convex.hull")
```

<!-- CHECK HERE: Gilles, when export_shp = TRUE the element $results was empty -->

--> *IT WORKS JUST FINE WIH ME, hope this is is not a problem linked to package or R version....*

```{r eoo2, eval = TRUE, collapse = TRUE}
EOO.hull <- EOO.computing(XY = occs[, c(1:3)], method.range = "convex.hull", export_shp = T)
EOO.hull$results
EOO.hull$spatial
```


### Area of occupancy (AOO)
```{r aoo, eval = TRUE, collapse = TRUE}
AOO <- AOO.computing(occs[, c(1:3)], 
                     cell_size_AOO = 2, nbe.rep.rast.AOO = 30)
```

### Number of subpopulations
```{r subpops, eval = TRUE, collapse = TRUE}
radius <- subpop.radius(XY = occs[, c(1:3)], 
                       factor.div = 10, quant.max = 0.9)
sub <- subpop.comp(XY = occs[, c(1:3)],
                   Resol_sub_pop = radius[,c("tax", "radius")])
```

<!-- CHECK HERE: Gilles, subpop.radius is returning an error: "some columns are not in the data.table: tax" -->
<!-- CHECK HERE: without the object radius I coul not test subpop.comp() -->

--> *CORRECTED !!*

### Number of locations

```{r locations, eval = TRUE, collapse = TRUE}
locs <- locations.comp(occs[, c(1:3),],  
                       method = "fixed_grid",
                            nbe_rep = 30,
                            cell_size_locations = 10, 
                            Rel_cell_size = 0.05, 
                            #threat_list = strict.ucs.spdf, 
                            #ID_shape = "NAME",
                            method_polygons = "no_more_than_one")
```

<!-- CHECK HERE: how can we include the shape file without the argument 'protec.areas' ?? -->

--> *CHEK parameters corrected above, it should work*


### Severe fragmentation
```{r fragmentation, eval = TRUE, collapse = TRUE}
merged_data <- merge(occs, radius, by = "tax", all.x = TRUE, sort = FALSE)
merged_data <- merge(merged_data, AOO[,c("tax", "aoo")], by = "tax", all.x = TRUE, sort = FALSE)
merged_data <- merged_data[, c("ddlat","ddlon","tax","radius","aoo")]
list_data <- coord.check(XY = merged_data,
                              listing = TRUE)
fator = 1
for (i in seq_lentgth(list_data)) {
get.patches(list_data[[i]],
      cell_size = 2,
      nbe_rep = 0,
      AOO = as.double(unique(list_data[[i]]$AOO)),
      Resol_sub_pop = as.double(unique(list_data[[i]]$radius)),
      dist_isolated = as.double(unique(list_data[[i]]$radius)) * fator,
    )
}
critB$sever.frag <- 100 * critB$Nbe_subPop/ (critB$AOO / 4) > 50

```

--> *get.patches function is tagged as internal and not exported Should we export it and if yes, make it work for multi taxa and thus not necessitates the loop ???*

--> *All table from EOO, AOO and locations computations now include a column 'tax' which simplify merging*

<!-- CHECK HERE: without the object radius I could not test coord.check() subpop.comp() -->



### Area of habitat (AOH)

```{r area-of-habitat, eval = TRUE, collapse = TRUE}
toto <- read.csv("data/ESACCI-LC-Legend.csv", as.is=TRUE)
hab.class <- toto$NB_LAB[grepl("ForestCover", toto$LegendTreeCoSimp)]

## Loading aggregated rasters and extracting the habitat loss/quality
ano1 = 2000
hab.map <- raster::stack(paste0("data/ESA_Land_Cover_map_", ano1, "_2015_AF_1km.tif"))
names(hab.map) <- paste("ESA",c(ano1,2015), sep=".")
anos <- 2015 - ano1 

toto.1 <- EOO.habitat(EOO.poly, 
                      hab.map, 
                      hab.class = hab.class, 
                      years = anos,
                      parallel = TRUE, NbeCores = 5)
```


<br/><br/>

## Small and declining populations (Criterion C) 


```{r criterion-C, eval = TRUE, collapse = TRUE}
critC <- criterion_C(x = mean.pop,
                       assess.year = 2018,
                       project = FALSE,
                       recent.year = 2000,
                       subcriteria = c("C1","C2"),
                       generation.time = gen.length,
                       prop.mature = p.mature,
                       subpop.size = NULL, # subpop.sizes,
                       correction = early.sucession)
```

## Species geographic range (IUCN criterion B)


```{r combine-B, eval = TRUE, collapse = TRUE}
# est.decline <- critC$cont.decline
# eoo.decline <- readRDS("data/EOO_hab_loss_2000_2015_uncropped.rds")
# declineB <- eoo.decline$rel.loss

```

```{r criterion-B, eval = TRUE, collapse = TRUE}
critB <- cat_criterion_b(EOO = EOO.hull$results,
                                  AOO = AOO,
                                  locations = locs$locations,
                                  sever.frag = sever.frag,
                                  decline = declineB,
                                  protected.threshold = 100)
```


## Population size reduction (IUCN criterion A)
                       
```{r criterion-A, eval = TRUE, collapse = TRUE}
critA <- criterion_A(mean.pop, assess.year = 2018,
                     project.years = NULL, subcriteria = c("A2"),
                     generation.time = gen.length,
                     exploitation = harvest)
```


## Very small population sizes (IUCN criterion D)

```{r criterion-D, eval = TRUE, collapse = TRUE}

```


# IUCN criteria assessments

## Combining all metrics
```{r all-criteria, eval = TRUE, collapse = TRUE}

```


## Near threatened

```{r near-threatened, eval = TRUE, collapse = TRUE}
near.threatened()
```



## Consensus assessment

```{r consensus, eval = TRUE, collapse = TRUE}
cat_mult_criteria()
```



