---
title: "Conservation assessments based on multiple-IUCN criteria using ConR"
# date: "`r format(Sys.time(), '%B %Y')`"
# author: 
#   - Renato A. F. de Lima^[Universidade de São Paulo, https://github.com/LimaRAF] # and Gilles Dauby^[IRD, https://github.com/gdauby]
output:
    rmarkdown::pdf_document:
        toc: true
        number_sections: true
    md_document:
      variant: gfm
vignette: >
  %\VignetteIndexEntry{Introduction to ConR}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEngine{rmarkdown::render}
    \usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", echo = TRUE)
```

```{r load, echo = FALSE, eval = TRUE, include=FALSE}
devtools::load_all() # for development
library(sf)
```

<br/><br/>

This vignette shows how to use ``ConR`` to apply a workflow for conducting
conservation assessments for multiple species using the IUCN criteria A, B, C
and D. Here, we use only three species to make it simple. But ``ConR`` was
developed to perform assessments for many more species at the same time.

<br/><br/>

# Definitions

We will use as examples three emblematic and threatened arborescent species that
are endemic or near-endemic to the Atlantic Forest in South America: *Araucaria angustifolia*, 
*Euterpe edulis* and *Paubrasilia echinata*. Thus, it is
necessary to obtain information for this specific region of the world (e.g.
habitat cover maps, protected areas), which should be available to run this 
tutorial.

```{r species, eval = TRUE, collapse = TRUE}
spp <- c("Araucaria angustifolia", "Euterpe edulis", "Paubrasilia echinata")
```

Another more general but important piece of information is the year of the
assessment itself (which depends on the information available) and the starting
year used to assess recent continuing decline (which depends on the region):

```{r year, eval = TRUE, collapse = TRUE}
assess.year <- 2018
recent.year <- 2000
```
  
<br/><br/>

# Species information

To perform the IUCN conservation assessments, we need different types of
species information related to their biology, distribution, population size,
threats, among other information.

## Occurrence records 

We will use real occurrence data for our target species. This example dataset
data is provided with ``ConR`` within the internal object `example_tutorial`,
but users can load a file containing a similar format. We will name it `occs` 
for simplicity.

```{r occurrences, eval = TRUE, collapse = TRUE}
data(example_tutorial)
occs <- example_tutorial$occurrences
head(occs, 2)
```

## Species biology and uses

Then, we need to define objects with the basic species information that are
necessary for the IUCN assessments.

```{r species-info, eval = TRUE, collapse = TRUE}
gen.length <- c(40, 65, 50) ## generation length in years
p.mature <- c(0.4485, 0.3028, 0.4143) ## proportion of mature individuals in the population
early.sucession <- c(0.9, 1, 1) ## correction applied to population decline
harvest <- c(10, 10, 10) ##  levels of exploitation causing reductions in population size (in %)
```

## Population sizes

To assess the IUCN criteria A, C and D, we also need to provide `ConR` the
population size information. For this tutorial, this information is also
provided within the internal object `example_tutorial`. We will name it
`pop.sizes`:

```{r pop-sizes, eval = TRUE, collapse = TRUE}
pop.sizes <- example_tutorial$population.sizes
years <- names(pop.sizes[,-1])
pop.sizes[, 1:6]
```

<br/><br/>

# Population metrics 

Before we start calculating the population metrics, we need to load ``ConR`` to
make the functions available in our R workspace.

```{r setup}
library(ConR)
```


## Population size decline

To match the time frame of 1-3 generation times into the past or into the
future, we need to estimate the population sizes for the years that we do not
have information available. These estimates are produced using the fit of
different statistical models (i.e. quadratic, exponential, logistic, general
logistic) to the population size trends. Here, we will only
estimate population sizes into the past, but one can also use the same `ConR`
function `pop.decline()` to estimate them into the future.


```{r est-pop-sizes, eval = TRUE, collapse = TRUE}
# Vector of years missing from the input data
possible.years <- 
  sort(unique(gen.length * rep(c(1:3), each=length(gen.length))), decreasing = TRUE)
missing.years <- assess.year - possible.years

# Estimating population sizes for the missing years
est.pop.sizes <- pop.decline(pop.sizes[,-1], taxa = spp, output = "predictions",
                             show_progress = FALSE)

# Re-organizing the information for downstream analyses
new.years <- est.pop.sizes[[1]][[1]]$years
ncols <- length(new.years)
nrows <- length(spp)
mean.pop <- 
  matrix(NA, ncol = ncols, nrow = nrows, dimnames = list(spp, new.years))

for(i in seq_along(spp)) mean.pop[i,] <- est.pop.sizes[[1]][[i]]$predicted
mean.pop <- cbind.data.frame(tax = spp, mean.pop, row.names = NULL)
```

We now have a new object called `mean.pop` which stores both the input
population sizes and the estimates based on the statistical models that best
describe the input data.


## Number of spatially valid and unique occurrences

We can begin by expecting how many valid occurrences are available for each
species.

```{r number-occs, eval = TRUE, collapse = TRUE}
nb.occs <- coord.check(XY = occs)
nb.occs <- nb.occs$unique_occs
```


## Extent of Occurrence (EOO)

We then use the species occurrences to compute their extent of occurrences (EOO)
in square kilometers using the function `EOO.computing()` and the
IUCN-recommended Convex Hull method.

```{r eoo, eval = TRUE, collapse = TRUE}
EOO.hull <- EOO.computing(XY = occs[, c(1:3)], method.range = "convex.hull", 
                          export_shp = TRUE, show_progress = FALSE)
```


## Area of occupancy (AOO)

Another important metric for IUCN assessments is the species' area of occupancy
(AOO), also in square kilometers, which is computed using the function
`AOO.computing()`. Here, we also set as defaults the IUCN recommendation of
using 2x2 km grid cells to compute AOO. `ConR` allows for computing AOO using 
multiple random starting positions of the grid, to make sure the AOO estimate 
is not influenced by the grid start position.


```{r aoo, eval = TRUE, collapse = TRUE}
AOO <- AOO.computing(occs[, c(1:3)], 
                     cell_size_AOO = 2, nbe.rep.rast.AOO = 30,
                     show_progress = FALSE)
```


## Number of subpopulations

The number of subpopulations for each species is also an important
information. The delimitation of subpopulations depends on species dispersal
abilities. If this information is available, it can be directly provided to the
function `subpop.comp()`. If it is not available, `ConR` offers the possibility
of estimating the maximum distance between species occurrences using the
function `subpop.radius()` and the circular buffer method. This distance can be
used as a proxy of species' dispersal abilities:

```{r subpops, eval = TRUE, collapse = TRUE}
radius <- subpop.radius(XY = occs[, c(1:3)], 
                       quant.max = 0.9)
sub <- subpop.comp(XY = occs[, c(1:3)],
                   resol_sub_pop = radius[,c("tax", "radius")],
                   show_progress = FALSE)
```

## Number of locations

One of the conditions to detect threat under the IUCN criterion B (restricted
geographic range) is related to the number of locations where the species occur.
Locations depend on the spatial extent of the threats faced by the species. Here
we used an extent of 10 km that makes sense for the target region, the Atlantic
Forest. This computation is performed using the function `locations.comp()`.

```{r locations, eval = TRUE, collapse = TRUE}
locs <- locations.comp(occs[, c(1:3),],  
                       method = "fixed_grid",
                       nbe_rep = 30,
                       cell_size_locations = 10, 
                       rel_cell_size = 0.05, 
                       #threat_list = strict.ucs.spdf, 
                       #id_shape = "NAME",
                       method_polygons = "no_more_than_one", 
                       show_progress = FALSE)
```
<!-- CHECK HERE: I managed to include the shape file but i am getting an error in line 618: do.call("rbind", shapes_loc)  ?? -->

## Severe fragmentation

Another condition to detect threat under the IUCN criterion B is whether the species populations is severely fragmented: _more than half of species AOO corresponds to habitat patches separated from others by a large distance_. Thus, to assess the level of fragmentation we need the information on the species AOO and the dispersal ability. The function ``severe_frag()`` computes the percentage of AOO that is isolated across a large distance.


```{r fragmentation, eval = TRUE, collapse = TRUE}
sever.frag <- severe_frag(XY = occs, 
                          resol_sub_pop = radius[,c("tax", "radius")], 
                          dist_isolated = radius$radius,
                          show_progress = FALSE)

```


## Area of habitat (AOH)

<!-- change the vignette text and code here after function AOH.estimation() is
ready and returning a data frame with the columns below:
AOH <- data.frame(tax = spp,
                  prop.EOO = c(18.80422, 18.74939, 16.89696), 
                  area.EOO = c(309588.7, 389226.0, 274717.6),
                  loss = c(1.587834, 1.529609, 1.535355), 
                  recover = c(1.616235, 1.494770, 1.317942), 
                  rel.loss = c(8.456802, 8.143053, 8.971141),
                  rate.loss  = c(0.5637868, 0.5428702, 0.5980761), 
                  rate.recover  = c(0.5730043, 0.5314911, 0.5199917)) -->

<!-- PREVIOUS COMMENTS/NOTES  -->
<!-- CHECK HERE: error. In the first version of this function 'years' was the
time interval between rasters (as described in the details of the function
documentation). Now I got this error: "the number of years provided is different
to the number of layers/polygon provided".
In addition, the function must return the relative change in AOH if multiple
raster are provide, so we can access decline for criterion B.
The previous columns returned by the function (which were used to assess
continuing decline) are in the data frame above -->

The area of habitat (AOH) is currently not an official population metric that is
compared against the IUCN Criteria. However, it provides important information
on the amount of habitat left for a given species within its EOO and more
importantly how this amount is changing through time.

`ConR` computes AOH using the function `AOH.estimation()`. If habitat maps are
available for different periods in time, the function can be used to infer the
continuing declines of the species AOH, which is an important condition to
assess the IUCN criterion B.

For this example, we do not provide habitat maps for the area of study (i.e.
rasters with the Atlantic Forest cover through time), which are demanding in
terms of storage and computational time. So we will assume that we already have
available a vector containing the relative losses (in %) of AOH previously
estimated for each species between 2000 and 2018.


```{r area-of-habitat, eval = TRUE, collapse = TRUE}
rel.loss <- c(8.5, 8.1, 9.0)
declineB <- ifelse(rel.loss >= 1, "Decreasing", "Not Decreasing")
```


<br/><br/>

# IUCN criteria assessments

## Population size reduction (IUCN criterion A)

The IUCN criterion A mainly evaluates paste and projects population size reductions. In `ConR`, it is assessed using the function `criterion_A()`, which requires population sizes through time (`mean.pop`) and generation lengths (`gen.length`). For this tutorial, we will assess only the subcriterion A2, which makes more sense in the specific context of the Atlantic Forest. 

`ConR` also allows to include information on exploitation levels for commercial species (e.g. timber), related to the base d of criterion A. These levels are provided using the argument `exploitation`.
                       
```{r criterion-A, eval = TRUE, collapse = TRUE}
critA <- criterion_A(x = mean.pop, 
                     assess.year = assess.year,
                     project.years = NULL, 
                     subcriteria = c("A2"),
                     generation.time = gen.length,
                     exploitation = harvest, show_progress = FALSE)
```

## Species geographic range (IUCN criterion B)

The function `criterion_B()` help to obtain assessments based on the criterion
B, as well as others parameters documenting the species geographic range.

```{r metrics-B, eval = TRUE, collapse = TRUE}
critB <- criterion_B(x = occs, 
                     AOO = AOO, 
                     EOO = EOO.hull$results, 
                     locations = locs$locations,
                     severe.frag = sever.frag,
                     subpops = sub, 
                     decline = declineB, show_progress = FALSE)
```


## Small and declining populations (IUCN criterion C) 

The IUCN criterion C is about small and declining populations. In `ConR`, this criterion is assessed using the function `criterion_C()`, which requires subpopulation sizes to apply the sub-criterion C2.

We do not have estimates of subpopulation sizes available. So, we take a simple
approach which is assuming that all subpopulations have the same size. And we store this simplification in the object `subpop.sizes`.


```{r subpop-sizes, eval = TRUE, collapse = TRUE}
id.assess <- which(names(mean.pop) %in% assess.year)
df <- merge(data.frame(tax = spp, '2018' = mean.pop[,id.assess]), sub)
subpop.sizes <- vector("list", dim(sub)[1])
names(subpop.sizes) <- sub$tax
for(i in seq_along(subpop.sizes)) {
  subpop.sizes[[i]] <- rep(df$X2018[i]/df$subpop[i], df$subpop[i])
}
```

Now, we apply criterion C itself using `criterion_C()`, which requires
information on population sizes through time (`mean.pop`), generation lengths
(`gen.length`) and proportion of mature individuals (`p.mature`), besides the
subpopulation sizes obtained above (`subpop.sizes`).

```{r criterion-C, eval = TRUE, collapse = TRUE}
critC <- criterion_C(x = mean.pop,
                     assess.year = assess.year,
                     project = FALSE,
                     recent.year = recent.year,
                     subcriteria = c("C1","C2"),
                     generation.time = gen.length,
                     prop.mature = p.mature,
                     subpop.size = subpop.sizes,
                     correction = early.sucession, show_progress = FALSE)
```

If C1 is listed within the argument `subcriteria`, the function returns the
estimated continuing decline:

```{r continuing-decline, eval = TRUE, collapse = TRUE}
est.decline <- critC$cont.decline
```

## Very small population sizes (IUCN criterion D)

The IUCN criterion D is assessed using the function `criterion_D()`. We have no
spatially explicit information for future human activities or stochastic events
for the Atlantic Forest. Therefore, here we assess the subcriterion D and not D2.
However, function `criterion_D()` can be used to assess D2 only if the AOO and the 
number of locations are provided.

```{r criterion-D, eval = TRUE, collapse = TRUE}
# accounting for exploitation of commercial species
p.mature.explo <- p.mature - harvest/100
# population sizes at the year of assessment
pop.sizes.assess <- mean.pop[[18]]
# assessing criterion D
critD <- criterion_D(pop.size = pop.sizes.assess, 
                     name_sp = mean.pop[[1]], 
                     AOO = AOO$aoo,
                     n.Locs = locs$locations$locations,
                     prop.mature = p.mature.explo, 
                     subcriteria = c("D", "D2"),
                     AOO.threshold = 20, 
                     Loc.threshold = 5)

```

## Combining the results from all criteria

Because we are assessing multiple criteria for the same set of species, we need
to combine the results from all criteria. We try to avoid duplicated columns in the final object `all.crit`.

```{r all-criteria, eval = TRUE, collapse = TRUE}
all.crit <- merge(critA, critB, by = "tax", all = TRUE)

rm.dup.columns <- which(names(critC) %in% c("assessment.year", "assessment.period"))
all.crit <- merge(all.crit, critC[, -rm.dup.columns], by = "tax", all = TRUE, 
                  suffixes = c(".A",".C"))

rm.dup.columns <- which(names(critD) %in% c("AOO"))
all.crit <- merge(all.crit, critD[, -rm.dup.columns], by = "tax", all = TRUE)

all.crit <- all.crit[order(all.crit$tax),]
```


## Near threatened

Next, we try to tell apart the Least Concern (LC) from Near-Threatened (NT)
species using the function `near.threatened()`:


<!-- change "category_B" by "B1" and "B2" after fixing the output of criterion_B() -->

```{r near-threatened, eval = TRUE, collapse = TRUE}
subcriteria <- c("A2", "category_B", "C1", "C2", "D")
for(i in 1:length(subcriteria)) {
  all.crit[,subcriteria[i]] <- 
    near.threatened(cats = all.crit[,subcriteria[i]],
                    EOO = all.crit$EOO,
                    AOO = all.crit$AOO,
                    decline = all.crit$any.decline,
                    pop.reduction = all.crit$reduction_A12,
                    pop.size = all.crit$pop.size,
                    pop.size.low = NULL,
                    locations = all.crit$locations,
                    sever.frag = all.crit$severe_frag,
                    ext.fluct = NULL,
                    subpop = all.crit$subpop, 
                    subcriteria = subcriteria[i])
}
```


## Consensus assessment

Finally, we obtained the consensus categories for all four IUCN criteria using
the function `cat_mult_criteria()`. Although the criteria resulting in the most
restrictive category for each species prevails, the function returns the
evaluation result using other criteria.

```{r consensus, eval = TRUE, collapse = TRUE}
consensus <- cat_mult_criteria(all.crit[, c("tax", subcriteria)])
all.crit <- cbind.data.frame(all.crit, 
                             consensus[, c("category","main.criteria","aux.criteria")])
```

The object `all.crit` now contains all relevant population metrics and the
results of the IUCN conservation assessments using the Red List criteria A-D.

